---
title: "Effect Size and Cohen's d"
author: "djani"
date: today
format:
  html:
    theme: cosmo               # Use a clean Bootstrap theme
    toc: true                  # Table of contents on the side
    toc-depth: 2               # Up to level 2 headers in TOC
    number-sections: false      # Number section headings
    code-fold: false           # Show all code by default
    code-line-numbers: false   # Don't show line numbers
    highlight-style: github    # Syntax highlighting style

execute:
  echo: true                   # Show code chunks
  results: 'markup'            # Inline output next to code
  collapse: true               # Collapse output into the same block as code
  comment: "#>"                # Prefix for output lines (like in knitr)
  warning: false               # Suppress warnings in output
  message: false               # Suppress messages (e.g., from package loading)

editor: visual                 # Use Quarto visual editor by default
lang: en                       # Set the document language
fontsize: 1em     
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  # cache = TRUE,
  fig.retina = 2,
  fig.width = 6,
  fig.asp = 2 / 3,
  fig.show = "hold"
)
```

## Example experiment

Imagine we are studying the effect of a new anti-ADHD medication on patient's ability to focus. We design a controlled experiment and measure the focus time of individuals with and without the drug.

```{r}
library(tidyverse)

# focus time in min 
placebo   <- c(15, 16, 13, 15, 14, 16, 16, 17 , 13, 15, 14)
treatment <- c(19, 19, 16, 18, 18, 17, 19, 18 , 18, 16, 16)

# mean time 
mean_placebo    <- mean(placebo)
mean_treatment  <- mean(treatment)

# standard deviations
sd_placebo   <- sd(placebo)
sd_treatment <- sd(treatment)

pooled_sd <- (sd_placebo + sd_treatment)/2
```

Now we have a fictitious experiment where it seems that the drug leads to increase of focus time.

```{r}
# for visualisation only
df <- 
  data.frame(
    mean = c(mean_placebo, mean_treatment),
    sd   = c(sd_placebo, sd_treatment)
  ) 

rownames(df) <- c("placebo", "treatment")

# focus time
df
```

## Statistical significance

Lets see if the increase between placebo and the treatment is statistically significant assuming p \< 0.05.

```{r}
t.test(treatment, placebo, "greater")
```

Yes the `diff(c(mean_placebo, mean_treatment))` is around `r round(diff(c(mean_placebo, mean_treatment)))` minutes and is statistically significant because there is a miniscule (only `r round(100*t.test(placebo, treatment, "greater")$p.value, 4)` %) chance that those two sets (`placebo` and `treatment`) stem from the same underlying population (and this in reality have the same mean).

But what about the effect?

## Effect size using Cohen's d

### Theory

The effect size shows **how much** the treatment affects the people. How big is the change. The formulation of effect size is

$$
d = \frac{M_{\text{med}} - M_{\text{placebo}}}{s_{\text{pooled}}}
$$

If the difference in the nominator would be equal the standard deviation ($M_{med} - M_{placebo} = s_{pooled}$ ) then $d = \frac{s_{pooled}}{s_{pooled}} = 1$ and we woud say that on average the treatment leads **to one standard deviation increase** in effect. In literature 0.2 is weak effect, 0.6 is a moderate effect, and everything above 0.8 standard deviations is a very strong effect.

Lets calculate $d$ for our fictitious example.

```{r}
d <- (mean_treatment - mean_placebo)/pooled_sd
d

# using function in
library(lsr)
cohensD(treatment, placebo)
```

### Interpretation

The effect of this drug is `r round(cohensD(treatment, placebo), 3)` times the pooled standard deviation of the study group - a very strong effect on focus time. On average, the drug leads to an increase in focus time by about (`d x pooled_sd`) `r round(cohensD(treatment, placebo) * pooled_sd, 3)` minutes.

So while statistical significance tells us that we can be sure that the numbers are trully different and not just a random realization of the same underlying process (or population), the effect size informs us about the **magnitude** of the difference.
